---
title: "Exploring Environmental Parameters in The Northern Gulf of Alaska"
author: "Megan Smith"
date: "2024-12-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE,
                      fig.path = "../output/")
```

 Load Libraries 
```{r, echo=FALSE}
library(tidyverse)
library(here)
library(lubridate)
library(ggmap)

```

Read in data 
```{r, echo= FALSE,results="hide"}
stat_cords<-read_csv(here("data", "Sample_coordinates.csv"))
ctd<- read_csv(here("data","ctdspatialclean.csv"))
chl<- read_csv(here("data","chlspatialclean.csv"))
intchl<- read_csv(here("data","intchlspatialclean.csv"))
```
Exploring the data 
```{r, echo= FALSE, results='hide'}
glimpse(ctd)
glimpse(chl)
glimpse(intchl)
glimpse(stat_cords)
```


```{r, echo= FALSE, results="hide"}
spatial_order<- c("PWS", "inner", "middle", "outer", "slope") # order I want the spatial zones in 
ctd$spatial <- factor(ctd$spatial, levels = spatial_order)
custom_colors <- c( # colors for each year 
  "2015" = "#C27A89",
  "2016" = "darkseagreen",
  "2017" = "#C6B9E5",
  "2018" = "aquamarine3",
  "2019" = "coral3",
  "2020" = "burlywood3",
  "2021" = "darkkhaki",
  "2022"="darkslategrey")
spat_colors<-c( # custom colors for spatial zone 
  "PWS"= "#8E0F28",
  "inner"="#87BFCC",
  "middle"= "#3A8299",
  "outer"="#1E3F66",
  "slope"="darkgrey"
  
)
```


# The Northern Gulf of Alaska Spring Bloom 

- Spring Bloom: highest productivity
- Time: April-May
- Spatial Variation: Inner shelf zones have higher productivity and are influenced by fresh water discharge (snow melt), and wind mixing. The outer shelf and slope are relatively stable and have the lowest amounts of production
- Interannual Variation: 2015-2016 and 2019 were marine heat waves which are seen in temperature figures below
- Sampling: Each station is only sampled one time during the spring bloom, so this is a small snapshot of what is happening during the bloom

# Map of The Northern Gulf of Alaska Shelf

```{r, echo= FALSE}
stat_cords_clean <- stat_cords %>%
  mutate(long = as.numeric(long), lat = as.numeric(lat)) %>% # making sure lat and long are numeric
  filter(!is.na(long) & !is.na(lat)) # filtering out NAs


site <- data.frame(
  lon = -148.195623,  # approx long for the middle of the NGA
  lat = 59.194642     # approx long for the middle of the NGA
)

NGA <- get_map( # pulling the map of the area
  location = site, 
  zoom = 7,         # adjusted zoom level to show more of the Gulf
  maptype = "satellite"  # selecting satellite as the view
)
ggmap(NGA) + # creating the map 
  geom_point(data = stat_cords_clean, aes(x = long, y = lat, color = location), size = 4) + # selecting the data and making the points size 4
  labs(title="Spatial Distibution of The Northern Gulf of Alaska Shelf", # adding titles for labels
       x="Longitude", y="Latitude", 
       color= "Zone Classification") +
  scale_color_manual(values = spat_colors) + # using custom colors
  theme_minimal() + # selecting minimal theme
  theme(
    plot.title = element_text(hjust = 0.5, size = 14), #plot title size
    legend.position = "right", # putting legend on right
    plot.background = element_rect(fill = "#F2EDE6", color = NA),  # background of entire plot
    panel.background = element_rect(fill = "#F2EDE6", color = NA)  # background of the map area
  )
```

---

# Depth Profile of Spatial Zones

```{r, echo=FALSE}
stat_cords$location <- factor(stat_cords$location, levels = c("PWS", "inner", "middle", "outer", "slope"))

ggplot(stat_cords, aes(x = location, y = depth, group = 1, color = location)) + # plotting depth profile
  geom_line(linewidth = 1) +  # size of connecting lines 
  geom_point(size = 3) +  # point size
  scale_y_reverse(
    breaks = seq(0, max(stat_cords$depth), by = 250)  # y axis breaks every 250
  ) +  
  scale_color_manual(values = spat_colors) +  # custom spatial colors
  labs(
    x = "Spatial Zone", # renaming labels
    y = "Depth (m)",
    title = "Station Depths by Spatial Zone"
  ) +
  theme_minimal() +
  theme(
    axis.line = element_line(color = "darkgrey"), # axis colors
    axis.text.x = element_text(angle = 45, hjust = 1), # angle of axis text
    legend.position = "none"  # removing lengend
  )
```


---

# Temperature by Spatial Zone

```{r, echo= FALSE}
ggplot(ctd, aes(x = spatial, y = temp, fill = factor(year), group = interaction(spatial, year))) +
  geom_boxplot( #boxplot for temp data 
    linewidth = 0.4, # line size
    fatten = 1.5, # width size
    outlier.size = 0.25 # size of outlier dots
  ) +
  scale_fill_manual(values = custom_colors) + # using custom colors
  labs(
    title = "2015-2022 Spring Bloom Temperature (°C) by Spatial Zone", # changing names of labels
    x = "Spatial Zone",
    y = "Temperature (°C)",
    fill = "Year"
  ) +
  theme_minimal() + # minimal theme
  theme(
    axis.line = element_line(color = "darkgrey"), # axis color to dark grey 
    axis.text.x = element_text(angle = 80, hjust = 1), # angle of axis text
    legend.position = "top" # placing lengend on top 
  ) +
  guides(
    fill = guide_legend( #selecting how I want the legend on top to be so its just squares
      nrow = 1, 
      title = "Year",
      override.aes = list(
        linetype = 0,
        linewidth = 0,
        shape = 22, # this make rectangles
        size = 5 # size of rectangle
      )
    )
  )


```

---

# Temperature by Year

```{r, echo=FALSE}
ggplot(ctd, aes(x = factor(year), y = temp, fill = spatial, group = interaction(spatial, year))) +
  geom_boxplot( # selecting boxplot
    linewidth = 0.4, # line width size
    fatten = 1.5, # width of box
    outlier.size = 0.25 # outlier dot size
  ) +
  scale_fill_manual(values = spat_colors) +  # using custom colors for spatial zones
  labs(
    title = "2015-2022 Spring Bloom Temperature (°C) by Year", # changing label names
    x = "Year",
    y = "Temperature (°C)",
    fill = "Spatial Zone"
  ) +
  theme_minimal() +
  theme(
    axis.line = element_line(color = "darkgrey"), # axis line grey 
    axis.text.x = element_text(angle = 80, hjust = 1), # changing angle of axis text
    legend.position = "top" # putting legend on top 
  ) +
  guides(
    fill = guide_legend( # making rectangles instead of the box shape
      nrow = 1, # only one row
      title = "Spatial Zone",
      override.aes = list(
        linetype = 0,
        linewidth = 0,
        shape = 22, # this is rectangles
        size = 5 # size of rectangles 
      )
    )
  )

```

# Now moving to looking at Chlorophyll α throughout the spatial zones
- This data is only from the surface to 50m where the most primary production is happening

# Total Chlorophyll α (μg/L) by Spatial Zone

```{r, echo= FALSE}
ggplot(chl, aes(x = factor(year), y = chltot, fill = spatial, group = interaction(spatial, year))) +
  geom_boxplot( # selecting boxplot for this data 
    linewidth = 0.4, #boxplot line witdth
    fatten = 1.5, # width of boxplot
    outlier.size = 0.25 # outlier dot size
  ) +
  scale_fill_manual(values = spat_colors) +  #  custom colors for spatial zones
  labs(
    title = "2015-2022 Chlorophyll α (μg/L) by Spatial Zone", # label names 
    x = "Year",
    y = "Chlorophyll α (μg/L)",
    fill = "Spatial Zone"
  ) +
  theme_minimal() +
  theme(
    axis.line = element_line(color = "darkgrey"), # axis line color
    axis.text.x = element_text(angle = 80, hjust = 1), # angle of axis text
    legend.position = "top" # legend on top
  ) +
  guides(
    fill = guide_legend( # creating rectangles instead of boxplots for legend 
      nrow = 1,
      title = "Spatial Zone",
      override.aes = list(
        linetype = 0,
        linewidth = 0,
        shape = 22, # rectangle shape
        size = 5 #size of rectangle
      )
    )
  )

```

---

# Integrated Chlorophyll α (mg m²) by Spatial Zone

```{r, echo= FALSE}
ggplot(intchl, aes(x = year, y = chl_int_total, color = spatial, group = spatial)) + 
  geom_line(linewidth = 1) +  # changing line width for readability
  geom_point(size = 3) +  # point size
  scale_color_manual(values = spat_colors) +  # sustom colors for each spatial zone
  scale_x_continuous(breaks = seq(min(intchl$year), max(intchl$year), by = 1)) +  # making xaxis ticks for each year
  labs(
    title = "2015-2022 Integrated Chlorophyll α (mg m²) by Spatial Zone", # adding title and creating custom titles for labs
    x = "Year",
    y = "Int. Chlorophyll α (mg m²)",
    color = "Spatial Zone"
  ) +
  theme_minimal() + # selecting minimal theme
  theme(
    axis.line = element_line(color = "darkgrey"), # choosing a black axis lines
    axis.text.x = element_text(angle = 45, hjust = 1), # axis text lines
    legend.position = "top" # moving legend to the top 
  ) +
  guides(
    color = guide_legend( # customizing the legend on top
      nrow = 1, # keeping it one row
      title = "Spatial Zone",
      override.aes = list( 
        linetype = 1, # makes the thicker line
        shape = 16, # this is a circle
        size = 5 # bigger size 
      )
    )
  )
```

---

### Now that we have seem the interannual trends, we can explore the yearly trends with [this Shiny App](https://meg-n-smith.shinyapps.io/NGA_PWS/)

 
