---
title: "Final Project: Phase one cleaning"
author: "Megan Smith"
date: "2024-12-07"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE,
                      fig.path = "../output/")
```

This is the first level of cleaning the data in order to make it into a shiny app. 

 Accompanying data dictionary will have all information and links to original data sets. 


# Loading in libraries 
```{r}
library(tidyverse)
library(here)
```

# Reading in data
```{r}

ctd<- read_csv(here("data","CTDfullcleanshareable.csv" ))
chl<- read_csv(here("data", "CHLfullcleanshareable.csv"))
intchl<- read_csv(here("data","CHLallIntegratedWide.csv"))
```

```{r echo= TRUE, results="hide"}

glimpse(chl)
glimpse(intchl)
glimpse(ctd)
```
Getting unique station names and ensuring they are the same throughout 
```{r, echo= TRUE, results= "hide"}
unique(ctd$station)
unique(chl$station)
unique(intchl$station)
```
# Begining to clean 
```{r}
stations <- c("GAK01", "PWS2", "GAK04", "GAK05", "GAK09","GAK13","GAK14","GAK15") # these are the only stations I want to keep
ctdpro <- ctd %>% 
  filter(station %in% stations) # filtering above stations
```
# Checking which stations are in each year
```{r, echo= TRUE, results= 'hide'}
stations_year <- ctdpro %>% 
  group_by(year) %>% 
  summarise(statyear = list(unique(station))) # making a list of the stations for each year
library(dplyr)


statdepths <- ctdpro %>%
  group_by(station) %>%        
  summarize(max_depth = max(depth, na.rm = TRUE))  # getting the maximum 'depth' for each 'station'


print(statdepths)

```
# Adding a spatial identifier: PWS, inner shelf, middle shelf, outer shelf, and slope
```{r}
ctdprospat <- ctdpro %>% 
  mutate(spatial = case_when( #adding a column to give each station a spatial identifier
    station == "PWS2" ~ "PWS",
    station == "GAK01" ~ "inner",
    station %in% c("GAK04", "GAK05") ~ "middle",
    station == "GAK09" ~ "outer",
    station %in% c("GAK13", "GAK14", "GAK15") ~ "slope", 
    TRUE ~ NA_character_  
  ))
```

# Selecting out the columns needed and saving data 
```{r}
ctdproclean<-ctdprospat %>% 
  select(stat_name, station, year, depth, temp, sal, dens, spatial) %>% 
  write_csv(here("data","ctdspatialclean.csv"))
```

# Repeating above for chl data 
```{r}
chlpro <- chl %>% 
  filter(station %in% stations)

chlprospat <- chlpro %>% 
  mutate(spatial = case_when(
    station == "PWS2" ~ "PWS",
    station == "GAK01" ~ "inner",
    station %in% c("GAK04", "GAK05") ~ "middle",
    station == "GAK09" ~ "outer",
    station %in% c("GAK13", "GAK14", "GAK15") ~ "slope",
    TRUE ~ NA_character_  
  ))
chlproclean<-chlprospat %>% # different columns within this data set
  select(stat_name, station, year, depth, chlsml, chllrg, chltot,spatial) %>% 
  write_csv(here("data","chlspatialclean.csv"))
```

# Repeating above for integrated chl data (note the integration was done previously using the above chl data)
```{r}
intchlpro <- intchl %>% 
  filter(station %in% stations)

intchlprospat <- intchlpro %>% 
  mutate(spatial = case_when(
    station == "PWS2" ~ "PWS",
    station == "GAK01" ~ "inner",
    station %in% c("GAK04", "GAK05") ~ "middle",
    station == "GAK09" ~ "outer",
    station %in% c("GAK13", "GAK14", "GAK15") ~ "slope",
    TRUE ~ NA_character_  
  ))
intchlproclean<-intchlprospat %>% 
  select(stat_name, station, year, chlsml_int, chllrg_int, chl_int_total,spatial) %>% # slightly different column names, but all three data sets are generally the same 
  write_csv(here("data","intchlspatialclean.csv"))
```

